Beta Diversity through Time and Space

### 2021-02-18 code/beta_diversity.Rmd
revised ms1

# Binomial smooth function (thanks, Hadley): https://ggplot2.tidyverse.org/reference/geom_smooth.html
binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "binomial"), ...)
}

# Site_pair prep function
# site_pair_prep_net <- function(dat, y) {
#   dat %>%
#     ungroup() %>%
#     filter(year1 == year2 & 
#              year1 == y & 
#              site1 != site2) %>% # I don't think we want to compare samples through within sites, introduces a new form of non-independence.
#     filter(metric == "WN") %>% # we'll focus for now on WN, but this could be done for partitions, too
#     mutate(weights = rep(1, dim(.)[1])) %>% # right now this analysis is unweighted; could weight by abundance, etc, if we wanted to
#     dplyr::select(distance = value, weights,
#                   s1.xCoord = x1, s1.yCoord = y1, 
#                   s2.xCoord = x2, s2.yCoord = y2, 
#                   s1.elev.mean = elev1, s1.yday = yday1,
#                   s2.elev.mean = elev2, s2.yday = yday2) %>%
#     data.frame()
# }

@@

## wrapper for extracting splines and prepping tibble for plotting
spline_prep <- function(x, y) {
  isplineExtract(x) %>%
    as.data.frame() %>%
    tibble() %>%
    mutate(x.Geographic = x.Geographic/1000) %>%
    pivot_longer(cols = c(x.Geographic, x.elev.mean), 
                 names_to = "x.var", values_to = "predictor", 
                 names_prefix = "x.") %>%
    pivot_longer(cols = c(y.Geographic, y.elev.mean), 
                 names_to = "y.var", values_to = "response", 
                 names_prefix = "y.") %>%
    filter(x.var == y.var) %>%
    mutate() %>%
    dplyr::select(-y.var)
}

@@

# Join into single data frame for visualization
species_beta <- full_join(bb_beta, fl_beta) %>%
  left_join(site_data, by = c("site1" = "site")) %>% # add site data
  dplyr::select(site1, site2, FL, BB, # first for site 1
                elev1 = elev.mean, x1 = x, y1 = y) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, site2, FL, BB, # then for site 2
                elev1, elev2 = elev.mean, x1, y1, x2 = x, y2 = y) %>%
  mutate(site1 = factor(site1), 
         site2 = factor(site2),
         elev.diff = elev1 - elev2,
         mean.elev = (elev1 + elev2)/2) %>%
  pivot_longer(cols = c(BB, FL), names_to = "group", values_to = "beta") %>%
  filter(elev.diff >= 0) # For each site pair, one will be higher and the other lower; thus, when their order is reversed only the sign of the difference changes. Since no two sites have exactly the same elevation, this deduplicates the data by dropping the ordering of a given pair of sites (one of the two orderings) that yields a negative elevation difference while retaining the alternative ordering that yields a positive elevation difference. By >= 0, I retain comparisons of a site to itself. I drop (in the ggplot call) within-site comparisons when I plot against elevation difference, and I isolate within-site comparison when I plot against time. In the latter case, it is also necessary to deduplicate, since each pair of site dates occurs in A-B and B-A. For that, I filter (again, in the ggplot call) so that yday.diff > 0, similarly dropping the orderings that yield negative differences. Note that all this deduplication nonsense is not necessary for the betalinkr call because it deduplicates internally.
```

@@

p <- c("OS" = "#e7298a", "ST" = "#d95f02", 
       "ST.h" = "#7570b3", "ST.l" = "#1b9e77", 
       "ST.lh" = "#66a61e", "WN" = "black")

# BB + FL elevation diff
g1 <- ggplot(filter(species_beta, 
                    site1 != site2), # only site-dates within 5 days
             aes(elev.diff, beta, color = group)) +
  geom_point(alpha = 0.25, pch = 16) +
  #binomial_smooth(formula = y ~ splines::ns(x, 3)) +
  geom_smooth(method = "lm", se = FALSE) +
  #facet_wrap(~group, ncol = 1) +
  xlab("Elevation difference (m)") +
  ylab("β-diversity") +
  theme_light()

# BB * FL elevation diff
g2 <- ggplot(filter(beta_mit_data, 
                      site1 != site2 & # no within-site comparisons
                      metric != "S" & # drop this metric
                      !metric.class %in% c("OS.partition", "WN.partition")), 
             aes(elev.diff, value, color = metric))  +
  geom_point(alpha = 0.20, pch = 16) +
  binomial_smooth(formula = y ~ splines::ns(x, 3), se = FALSE) +
  scale_color_manual(values = p, name = "partition") +
  xlab("Elevation difference (m)") +
  ylab(NULL) +
  theme_light()

# Patchwork plot for comparing species beta-diversity with interaction beta-diversity
png("../output/beta_stack.png", width = 6.5, height = 4.5, units = "in", res = 300)
(g1 | g2) + 
  plot_annotation(tag_levels = c("A", "B")) + 
  #plot_layout(guides = "collect") & 
  theme_light() 
dev.off()
```

>>>>

# # Binomial smooth function (thanks, Hadley): https://ggplot2.tidyverse.org/reference/geom_smooth.html
# binomial_smooth <- function(...) {
#   geom_smooth(method = "glm", method.args = list(family = "binomial"), ...)
# }

@@

@@

# Join into single data frame for visualization
species_beta <- full_join(bb_beta, fl_beta) %>%
  left_join(site_data, by = c("site1" = "site")) %>% # add site data
  dplyr::select(site1, site2, FL, BB, # first for site 1
                elev1 = elev.mean, x1 = x, y1 = y) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, site2, FL, BB, # then for site 2
                elev1, elev2 = elev.mean, x1, y1, x2 = x, y2 = y) %>%
  mutate(site1 = factor(site1), 
         site2 = factor(site2),
         elev.diff = elev1 - elev2,
         mean.elev = (elev1 + elev2)/2) %>%
  pivot_longer(cols = c(BB, FL), names_to = "group", values_to = "beta") %>%
  filter(elev.diff >= 0) # For each site pair, one will be higher and the other lower; thus, when their order is reversed only the sign of the difference changes. Since no two sites have exactly the same elevation, this deduplicates the data by dropping the ordering of a given pair of sites (one of the two orderings) that yields a negative elevation difference while retaining the alternative ordering that yields a positive elevation difference. By >= 0, I retain comparisons of a site to itself. I drop (in the ggplot call) within-site comparisons when I plot against elevation difference, and I isolate within-site comparison when I plot against time. In the latter case, it is also necessary to deduplicate, since each pair of site dates occurs in A-B and B-A. For that, I filter (again, in the ggplot call) so that yday.diff > 0, similarly dropping the orderings that yield negative differences. Note that all this deduplication nonsense is not necessary for the betalinkr call because it deduplicates internally.

# fl_beta_dist <- survey %>%
#   group_by(site, plant.sp) %>%
#   summarize(count = sum(flower.cover)) %>%
#   pivot_wider(names_from = plant.sp, values_from = count) %>%
#   replace(is.na(.), 0) %>%
#   column_to_rownames("site") %>%
#   vegdist(method = "jaccard", binary = TRUE, upper = TRUE) 
# 
# bb_beta_dist <- net %>%
#   group_by(site, bb.sp) %>%
#   summarize(count = n()) %>% 
#   pivot_wider(names_from = bb.sp, values_from = count) %>%
#   replace(is.na(.), 0) %>%
#   column_to_rownames("site") %>%
#   vegdist(method = "jaccard", binary = TRUE)
```

@@

p1 <- c("OS" = "#e7298a", "ST" = "#66a61e", "WN" = "black")

p2 <- c("ST.h" = "#7570b3", "ST.l" = "#1b9e77", "ST.lh" = "#d95f02")

# BB + FL elevation diff
g1 <- ggplot(filter(species_beta, 
                    site1 != site2), # only site-dates within 5 days
             aes(elev.diff, beta, color = group)) +
  geom_point(alpha = 0.25, pch = 16) +
  #binomial_smooth(formula = y ~ splines::ns(x, 3)) +
  #geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE) +
  #facet_wrap(~group, ncol = 1) +
  xlab("Elevation difference (m)") +
  ylab("β-diversity") +
  theme_light()

# BB * FL elevation diff (aggregate)
g2 <- ggplot(filter(beta_mit_data, 
                      site1 != site2 & # no within-site comparisons
                      metric != "S" & # drop this metric
                      metric.class == "aggregate"), 
             aes(elev.diff, value, color = metric))  +
  geom_point(alpha = 0.20, pch = 16) +
  #binomial_smooth(formula = y ~ splines::ns(x, 3), se = FALSE) +
  #geom_smooth(method = "gam", formula = y ~ s(x, k  = 4), se = FALSE) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE) +
  scale_color_manual(values = p1, name = "β partition") +
  xlab(NULL) +
  ylab("Prop. β-diversity") +
  theme_light()

# BB * FL elevation diff (ST partition)
g3 <- ggplot(filter(beta_mit_data, 
                      site1 != site2 & # no within-site comparisons
                      metric.class ==  "ST.partition"), 
             aes(elev.diff, value, color = metric))  +
  geom_point(alpha = 0.20, pch = 16) +
  #binomial_smooth(formula = y ~ splines::ns(x, 3), se = FALSE) +
  #geom_smooth(method = "gam", formula = y ~ s(x, k  = 4), se = FALSE) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE) +
  scale_color_manual(values = p2, name = "ST partition") +
  xlab("Elevation difference (m)") +
  ylab("Prop. species turnover") +
  theme_light()

# Patchwork plot for comparing species beta-diversity with interaction beta-diversity
png("../output/beta_stack.png", width = 6.5, height = 4.5, units = "in", res = 300)
(g1 | (g2 / g3)) + 
  plot_annotation(tag_levels = c("A", "B", "C")) + 
  #plot_layout(guides = "collect") & 
  theme_light() 
dev.off()
```

### MRM analysis
```{r}
library(ecodist)

survey_sample_diff <- survey %>%
  group_by(site) %>%
  summarize(samples = length(unique(date))) %>%
  column_to_rownames("site") %>%
  vegdist(method = "euclidean", upper = TRUE) %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "site1") %>%
  pivot_longer(cols = -site1, names_to = "site2", values_to = "sample.diff")

net_sample_diff <- net %>%
  group_by(site) %>%
  summarize(samples = length(unique(date))) %>%
  column_to_rownames("site") %>%
  vegdist(method = "euclidean", upper = TRUE) %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "site1") %>%
  pivot_longer(cols = -site1, names_to = "site2", values_to = "sample.diff")

mrm_survey_data <- species_beta %>%
  filter(group == "FL" & site1 != site2) %>%
  left_join(survey_sample_diff) %>%
  dplyr::select(site1, site2, beta, elev.diff, sample.diff) %>%
  mutate(elev.diff = scale(elev.diff), # gotta scale variables for MRM
         sample.diff = scale(sample.diff))

mrm_net_data <- species_beta %>%
  filter(group == "BB" & site1 != site2) %>%
  left_join(net_sample_diff)  %>%
  dplyr::select(site1, site2, beta, elev.diff, sample.diff) %>%
  mutate(elev.diff = scale(elev.diff),
         sample.diff = scale(sample.diff))

mrm_beta_data <- beta_mit_data %>%
  filter(site1 != site2) %>%
  left_join(net_sample_diff)  %>%
  dplyr::select(site1, site2, value, metric, elev.diff, sample.diff) %>%
  mutate(elev.diff = scale(elev.diff),
         sample.diff = scale(sample.diff))

mrm_fl <- MRM(beta ~ elev.diff + sample.diff, 
              data = mrm_survey_data,
              nperm = 1000,
              method = "logistic")

mrm_bb <- MRM(beta ~ elev.diff + sample.diff, 
              data = mrm_net_data,
              nperm = 1000,
              method = "logistic")

mrm_WN <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "WN"),
              nperm = 1000,
              method = "logistic")

mrm_ST <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "ST"),
              nperm = 1000,
              method = "logistic")

mrm_ST.l <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "ST.l"),
              nperm = 1000,
              method = "logistic")

mrm_ST.h <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "ST.h"),
              nperm = 1000,
              method = "logistic")

mrm_ST.lh <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "ST.lh"),
              nperm = 1000,
              method = "logistic")

mrm_OS <- MRM(value ~ elev.diff + sample.diff, 
              data = filter(mrm_beta_data, metric == "OS"),
              nperm = 1000,
              method = "logistic")

# Partial Mantel tests are another option, perhaps more routine to interpret
# mantel_fl <- mantel(beta ~ elev.diff + sample.diff, 
#               data = mrm_survey_data,
#               mrank = TRUE)
# 
# mantel_bb <- mantel(beta ~ elev.diff + sample.diff, 
#               data = mrm_net_data,
#               mrank = TRUE)
# 
# mantel_WN <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "WN"),
#                     mrank = TRUE)
# 
# mantel_ST <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "ST"),
#                     mrank = TRUE)
# 
# mantel_ST.l <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "ST.l"),
#                     mrank = TRUE)
# 
# mantel_ST.h <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "ST.h"),
#                     mrank = TRUE)
# 
# mantel_ST.lh <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "ST.lh"),
#                     mrank = TRUE)
# 
# mantel_OS <- mantel(value ~ elev.diff + sample.diff, 
#                     data = filter(mrm_beta_data, metric == "OS"),
#                     mrank = TRUE)
```
